% !TEX root = main.tex

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments} 

\thispagestyle{empty}

We would like to thank the people below for their feedback, suggestions and corrections
on this document.
It is understood that we, the authors, are to blame for any remaining errors
and omissions.


\section*{Version 2.0}

Version 2.0 of the SMT-LIB standard was developed with the input of 
the whole SMT community and
three international work groups consisting of developers and users of SMT tools:
the SMT-API work group, led by A.~Stump,
the SMT-LOGIC work group, led by C.~Tinelli, and
the SMT-MODELS work group, led by C.~Barrett.
The Version 2.0 document was written by C.~Barrett, A.~Stump and C.~Tinelli.

Particular thanks are due to the following work group members, 
who contributed numerous suggestions and helpful constructive criticism 
in person or in email discussions:
Nikolaj Bj{\o}rner, 
Sascha Boehme, 
David Cok, 
David Deharbe, 
Bruno Dutertre,
Pascal Fontaine,
Vijay Ganesh,
Alberto Griggio,
Jim Grundy,
Paul Jackson, 
Albert Oliveras, 
Sava Krsti\'c,
Michal Moskal, 
Leonardo de Moura,
Philipp R\"ummer,
Roberto Sebastiani,
and
Johannes Waldmann.

Thanks also go to
David Cok,
Morgan Deters,
Anders Franz\'en,
Amit Goel,
Jochen Hoenicke,
and
Tjark Weber
for additional feedback on the standard,
and to Jochen Hoenicke, Philipp R\"ummer,
and above all David Cok,
for their careful proof-reading of earlier versions of the Version 2.0 document.

\section*{Version 2.5}

Version 2.5 was developed again with the input of the SMT community, 
based on their experience with Version 2.0.

Special thanks go to the following people
for their thoughtful feedback and useful suggestions:
Martin Brain,
Adrien Champion,
J{\"u}rgen Christ,
David Cok,
Morgan Deters,
Bruno Dutertre,
Andrew Gacek,
Alberto Griggio,
Jochen Hoenicke,
Tim King,
Tianyi Liang,
Aina Niemetz,
Margus Veanes,
and
Christoph Wintersteiger.
We also thank David Cok and J\"{u}rgen Christ for their very careful 
proof reading of this document.


\section*{Version 2.6}

The main difference between 2.6 and Version 2.5 is the addition of algebraic datatypes to the language.
Special thanks go to the following people
for their feedback and useful suggestions on this extension:
Guillaume Bury,
David Cok,
Tim King,
Viktor Kuncak,
Jochen Hoenicke,
Pierre van de Laar,
Calvin Loncaric,
Andres N{\"o}tzli,
Andrew Reynolds,
and
Cristina Serban.

\section*{\thisversion}

\begin{newver}
The main difference between 2.7 and Version 2.6 is the inclusion of two extensions to
the SMT-LIB base logic:
$(i)$ support for prenex polymorphism in user-defined sorts and functions, and
$(ii)$ an extension to higher-order logic obtained, 
in essence, by the introduction of a theory of functions.
Aside from new functionality, which assigns special meaning 
to the symbols \ter{\_}, \ter{lambda}, and \ter{->},  
%\cbrem{What are they?  Either spell
%  them out or give a pointer to where they are described specifically.}
Version 2.7 is fully backward-compatible with Version 2.6
and is designed to ease the transition to the upcoming Version 3,
whose base logic will be a higher-order logic with dependent (and polymorphic)
types.
Special thanks go to 
Nikolaj Bj{\o}rner and
Andrew Reynolds for proposing this transitional version and 
for their feedback and useful suggestions.
We also thank Guillaume Bury and Levent Erkök for their feedback on the reference document and the standard.

Pascal Fontaine is grateful to Jasmin Blanchette for his support through his European Research Council (ERC) starting grant Matryoshka (713999).  
He also thanks Amazon for its support through the Amazon Research Award program 
(SMT: Modules, Formats, and Standards).
\end{newver}
